---
title: "Building ecological benchmarks using BUILDER"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{builder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Intro
BUILDER is a user friendly software application, developed in C# .NET framework, to assist conservation planners with the design of ecological benchmarks. This package provides a set of R functions to call BUILDER and fetch the main output table. The partner package `beaconstools` provides functions to process and analyse output tables from BUILDER in R.  

BUILDER explicitly incorporates hydrologic connectivity for the integration of aquatic and terrestrial conservation planning in protected areas design. BUILDER constructs ecological benchmarks using a deterministic construction algorithm that aggregates catchments to a user defined size and intactness. For the purpose of this demo, we will refer to the conservation areas created by BUILDER as benchmarks, however BUILDER can also be used to design conservation areas not intended to serve as benchmarks.  

BUILDER constructs ecological benchmarks based on three principal criteria:  

* Intactness, a measure of the absence of human industrial activity and a proxy for the intactness of biological and physical processes
* Hydrologic connectivity of intact water catchments, as a measure of the integrity of aquatic systems
* Size, as a measure of the resilience of the system to disturbance

## Catchments
The building blocks for BUILDER is a set of catchments with associated flow directions for the underlying stream network. This allows BUILDER to identify upstream and downstream catchments for any given location.  

## Aggregation algorithm
BUILDER ‘grows’ ecological benchmarks using a deterministic construction algorithm that aggregates catchments. This process enables intactness, size and hydrologic criteria to be incorporated into benchmark design.   

The starting catchments for building benchmarks are called seeds. Starting from a seed, BUILDER grows a benchmark by aggregating neighbouring catchments that exceed the user defined catchment level intactness threshold. As catchments are added, growth is prioritized in the upstream direction. This emphasizes inclusion of headwaters within benchmarks, and encourages the formation of benchmarks along stream networks. Further, by adding all eligible upstream catchments before re-querying for more, benchmarks are grown with a more compact shape. If re-querying were to occur after adding each upstream catchment, benchmarks would tend to grow along single streams, resulting in linear configurations. To further compact the shape of the benchmark, BUILDER prioritizes the addition of catchments within hydrological zones (i.e. Fundamental Drainage Areas, FDAs) associated with the benchmark before adding catchments in a neighbouring FDA.  

Once all available upstream catchments are added, BUILDER grows downstream. Downstream growth adds all neighbouring catchments downstream of the catchment aggregation then re-queries for upstream catchments. Benchmarks are grown until: 

  1. There are no more neighbouring, intact catchments to add, or 
  2. The benchmarks area exceeds the user-specified Area Target. 

If the latter occurs, and the area-weighted intactness of the benchmark exceeds the benchmark-level intactness threshold, then the benchmark is included in the output.

## R functions

**neighbours()**

Prepares a neighbours table listing pairwise catchment neighbours. Used as an input to `builder()`.

**seeds()**

Prepares a seeds table listing catchments to use as seeds when building benchmarks. Used as an input to `builder()`.

**builder()**

Prepares all input files and parameters and passes them to the BUILDER executable where each seed is built into a benchmark.


## BUILDER exe
The function `builder()` prepares the input files and parameters and passes them to a command line version of BUILDER executable software that resides inside the `benchmarkbuilder` package. The function then fetches and returns the main BUILDER output table into the R environment. 

Ultimately we hope to rewrite the BUILDER executable in R to avoid having to call a third party application. In the meantime, this package allows users to access and automate calls to BUILDER from R.

## Output
The BUILDER executable outputs a number of csv tables describing the catchments making up each benchmark, as well as information on hydrology and a range of summary statistics. There are two options for accessing this output in R:

#### Default behavior
The default behavior for `builder()` is for the BUILDER executable to save it's output csv tables to a temp folder. The main output table listing all catchments making up each benchmark is returned by the function in R and the temp folder is deleted.

This workflow works well when a user wants to complete a single BUILDER run and process the subsequent benchmarks using the various functions in the `beaconstools` package. This workflow is also the safest option because it avoids file path issues that can cause the BUILDER executable to crash if an invalid directory is provided.

#### out_dir
If an output folder directory is provided using the `out_dir` argument, the BUILDER executable will attempt to save the output files to this folder. The folder will not be deleted after the function runs, and the main output table listing all catchments making up each benchmark is still returned (this is the file name ending in 'COLUMN_All_Unique_BAs.csv'). This workflow gives the user access to all BUILDER executable output files. It can also be useful for automating multiple `builder()` calls and saving the output in different folders to be accessed later using the `beaconstools` package.

#### beaconstools
The output from `builder()` is a table listing all catchments associated with each benchmark. If an `out_dir` was provided, this table can also be accessed using `read.csv()` from the `out_dir` (file name ending in 'COLUMN_All_Unique_BAs.csv'). All subsequent processing, such as creating polygons of benchmarks and assessing ecological attributes is done using `beaconstools`. Below we demonstrate a simple example of running `builder()` and using `beaconstools` to generate shapefiles.

## Example: running `builder()` and creating polygons

```{r, echo=F}
# DELETE THIS WHEN CLOSING #14
builder_local_path <- "C:/Temp/BenchmarkBuilder_cmd.exe"
```

```{r message=FALSE, fig.width=7, fig.height=5}
library(benchmarkbuilder)

# generate neighbours table for catchments
nghbrs <- neighbours(builder_catchments_sample)

# Get seed list for all catchments with 100% intactness. Set benchmarks area target to 500km2.
seed <- seeds(catchments_sf = builder_catchments_sample,
             filter_intactness_col = "intact", filter_intactness_threshold = 1,
             areatarget_value = 500000000)

# Run builder() using the catchments, the seeds table with 500km2 area target, and the neighbours table
# Only build benchmarks with catchments that are 80% intact 
# Only return benchmarks with an area-weighted intactness of at least 95%
benchmarks_tab <- builder(catchments_sf = builder_catchments_sample, seeds = seed, neighbours = nghbrs,
                          catchment_level_intactness = 0.8, benchmark_level_intactness = 0.95)

# Note: the same table can be created using the out_dir option shown below.
# Note: the BUILDER output table has an OID column which is automatically removed when used in beaconstools
#
# benchmarks_dir <- file.path(tempdir(), "builder_output")
# builder(catchments_sf = builder_catchments_sample, seeds = seed, neighbours = nghbrs,
#                          catchment_level_intactness = 0.8, benchmark_level_intactness = 0.95,
#                          out_dir = benchmarks_dir)
# benchmarks_path <- list.files(benchmarks_dir, pattern = "COLUMN_All_Unique_BAs.csv", full.names = TRUE)
# benchmarks_tab <- read.csv(benchmarks_path)

# BUILDER created 10 benchmarks that met the area target using 80% intact catchments, and met the 95% 
# area-weighted intactness threshold
# View the output table listing the catchments making up the 166 returned benchmarks
head(benchmarks_tab)

# Open beaconstools
library(beaconstools)
library(dplyr)
library(sf)

# Convert benchmarks to polygons
benchmarks_poly <- dissolve_catchments_from_table(catchments_sf = builder_catchments_sample, 
                                             input_table = benchmarks_tab, 
                                             out_feature_id = "network")

# Make map of the catchments, colored green if intactness > 80%
# Add the outlines of all 166 benchmarks
# Highlight one benchmark in yellow as an example
plot(builder_catchments_sample$geometry[builder_catchments_sample$intact>=0.8], lty = 0, col = 'seagreen3')
plot(builder_catchments_sample$geometry[builder_catchments_sample$intact<0.8], lty = 0, col = 'grey', add = T)
plot(benchmarks_poly$geometry, add = T, border = 'darkgreen')
plot(benchmarks_poly$geometry[10], add = T, border = "gold", lwd = 2)

# Note how benchmarks only cover the green catchments areas because of the 80% catchment level intactness value used.
# Also note that not all patches of 80% intact catchments contain benchmarks. This is because the area and hydrology 
# requirements could not be met in those regions.

# Show the individual catchments within the example benchmark
plot(benchmarks_poly$geometry[10])
plot(builder_catchments_sample$geometry, add = T, col = 'seagreen3', border = 'grey40')
plot(benchmarks_poly$geometry[10], add = T, border = "gold", lwd = 2)

```

